---
title: "Exercise Two (Answers)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidybiology)
data(chromosome)

```

For this exercise, we will be using the `chromosome` data set from the `tidybiology` package

Q1. Take a look at the `chromosome` dataframe. What's a good function for this?
```{r}
glimpse(chromosome)
```

Q2. Make a density plot for the variable `length_mm`
```{r}
ggplot(chromosome, aes(length_mm)) +
  geom_density()
```

Q3. Add two vertical lines to this density plot that correspond to the mean and median of `length_mm` respectively
Hint: Use `geom_vline()` for this
```{r}
mean_length <- mean(chromosome$length_mm)
median_length <- median(chromosome$length_mm)

ggplot(chromosome, aes(length_mm)) +
  geom_density() +
  geom_vline(xintercept = mean_length) +
  geom_vline(xintercept = median_length)
```

Would you say this plot is skewed? Here's a nice chapter on the relationship between mean, median and skewness - https://openstax.org/books/introductory-business-statistics/pages/2-6-skewness-and-the-mean-median-and-mode

Q4. Remove the y-axis label from the plot above. Also, change the x-axis label to "Length (mm)" 
```{r}
ggplot(chromosome, aes(length_mm)) +
  geom_density() +
  geom_vline(xintercept = mean_length) +
  geom_vline(xintercept = median_length) +
  labs(x = "Length (mm)",
       y = NULL)
```

Q5. Make a scatter plot for `protein_codinggenes` versus `pseudo_genes`
BONUS: Using the function `geom_text_repel()` from the `ggrepel` package, label the point that corresponds to the chromosome with the highest number of protein coding genes (i.e. If it's chromosome 1, then the point will have the label "1")  

Note: You might have to install this package. You can find it on CRAN.
```{r}
library(ggrepel)

ggplot(chromosome, aes(protein_codinggenes, pseudo_genes)) +
  geom_point() +
  geom_text_repel(data = filter(chromosome, protein_codinggenes == max(protein_codinggenes)), aes(label = id)) # in this line, we're filtering on the fly (we can do that!) so that the resulting dataframe just has one row i.e. the one for the chromosome that has the highest number of protein coding genes. We then add a label by using the `label = ` parameter
# NOTE: `label = id` must go within the `aes()` function because we are mapping data from our dataframe onto the plot
```

