---
title: "Exercise Two"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidybiology)
data(chromosome)

```

For this exercise, we will be using the `chromosome` data set from the `tidybiology` package

Q1. Take a look at the `chromosome` dataframe. What's a good function for this?
```{r}

```

Q2. Make a density plot for the variable `length_mm`
```{r}

```

Q3. Add two vertical lines to this density plot that correspond to the mean and median of `length_mm` respectively
Hint: Use `geom_vline()` for this
```{r}
mean_length <- mean(chromosome$length_mm)
median_length <- median(chromosome$length_mm)


```

Would you say this plot is skewed? Here's a nice chapter on the relationship between mean, median and skewness - https://openstax.org/books/introductory-business-statistics/pages/2-6-skewness-and-the-mean-median-and-mode

Q4. Remove the y-axis label from the plot above. Also, change the x-axis label to "Length (mm)" 
```{r}

```

Q5. Make a scatter plot for `protein_codinggenes` versus `pseudo_genes`
BONUS: Using the function `geom_text_repel()` from the `ggrepel` package, label the point that corresponds to the chromosome with the highest number of protein coding genes (i.e. If it's chromosome 1, then the point will have the label "1")  

Note: You might have to install this package. You can find it on CRAN.
```{r}
library(ggrepel)


```

