---
title: "Exercise Three (Answers)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DESeq2)
library(tidyverse)
remotes::install_github("compgenomr/compGenomRData")
```

# Differential expression analysis in colorectal cancer

For this exercise, we will be using a subset of count data from a colorectal cancer study, as described here - 
http://compgenomr.github.io/book/gene-expression-analysis-using-high-throughput-sequencing-technologies.html#computing-different-normalization-schemes-in-r

First run all the code in the chunk below. This code prepares the `countData` and `colData` objects we will need to make a DESeqDataSet object   

```{r}
counts_file <- system.file("extdata/rna-seq/SRP029880.raw_counts.tsv",
                           package = "compGenomRData")
coldata_file <- system.file("extdata/rna-seq/SRP029880.colData.tsv",
                            package = "compGenomRData")

counts <- as.matrix(read.table(counts_file, header = T, sep = '\t'))

#remove the 'width' column
countData <- as.matrix(subset(counts, select = c(-width)))
#define the experimental setup 
colData <- read.table(coldata_file, header = T, sep = '\t', 
                      stringsAsFactors = TRUE)
```

Q1. Looking at `colData`, what type differential expression analysis could you reasonably perform? What would the design formula for this comparison be?
A1. We could compare gene expression between the groups `CASE` and `CTRL`
We would use the following design formula for this comparison - `~ group` 

Q2. How would make an appropriate DESeqDataSet object?
A2.
```{r}
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ group) 
```

Q3. How would you perform differential expression analysis? How would you inspect the results of this analysis?
A3.
```{r}
dds <- DESeq(dds)

results(dds)
```

Q4. In the analysis above, we ended up comparing log2FoldChange of CTRL vs CASE. How would you perform this analysis so that you were comparing CASE vs CTRL?
A4.
```{r}
results(dds,
        contrast = c("group", "CASE", "CTRL"))
```
