---
title: "Exercise Four (Answers)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DESeq2)
library(tidyverse)
#remotes::install_github("compgenomr/compGenomRData")
# uncomment the line above if you DON'T already have this package installed
```

For this exercise, we will be using a subset of count data from a colorectal cancer study, as described here - 
http://compgenomr.github.io/book/gene-expression-analysis-using-high-throughput-sequencing-technologies.html#computing-different-normalization-schemes-in-r

First run all the code in the chunk below. This code prepares the `countData` and `colData` objects we will need to make a DESeqDataSet object  

Run *all* of the code in the chunk below before attempting the subsequent questions

```{r}
counts_file <- system.file("extdata/rna-seq/SRP029880.raw_counts.tsv",
                           package = "compGenomRData")
coldata_file <- system.file("extdata/rna-seq/SRP029880.colData.tsv",
                            package = "compGenomRData")

counts <- as.matrix(read.table(counts_file, header = T, sep = '\t'))

#remove the 'width' column
countData <- as.matrix(subset(counts, select = c(-width)))
#define the experimental setup 
colData <- read.table(coldata_file, header = T, sep = '\t', 
                      stringsAsFactors = TRUE)

# run DESeq
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ group) 
dds <- DESeq(dds)
res <- results(dds,
           contrast = c("group", "CASE", "CTRL"))
```
Q1. How would you make a PCA plot using the `dds` object? Remember to first `rlog` transform this object
A1.
```{r}
rld <- 

```

Q2. What does the plot above tell you?
A2. There is good separation between the `CASE` and `CONTROL`. PC1 accounts for the large majority of total variance

Q3. Plot a heatmap of the ten most variable genes
A3.
```{r}
library("genefilter")

```

Q4. Retrieve the Ensembl IDs for the gene names in `res_df`
A4.
```{r}
library("biomaRt")
res_df <- as_tibble(res, rownames = "gene_name") 


```








