---
title: "Class Two: Computational Biology Reading Group"
author: "Your name"
date: "Oct 27, 2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) 
devtools::install_github("hirscheylab/tidybiology", force = TRUE)
library(tidybiology)

# we can import the required data in this chunk
golub_subjects <- readr::read_csv("https://raw.githubusercontent.com/BAREJAA/website_for_john/master/datasets/golub_kaggle/golub_subjects.csv")

data(proteins)
```

# Boxplots in base R

Make boxplots for all the columns in the data frame `golub_subjects` EXCEPT for the following columns - `gene_name`, `mean_ALL`, `mean_AML`  

Hints - 
- which function would you use to work out how many columns your data frame has?
- Using this information, select the relevant columns
- the `boxplot()` function can take a whole data frame as input. If you do this, this function will plot separate boxplots for each column

Step 1: Insert an R chunk below this line - 



# Exploring the `proteins` dataframe
```{r}
# how many rows and columns does `proteins` contain?


# introducing a more powerful function - glimpse()
glimpse(proteins)

# create a new data frame that only contains the following columns - uniprot_id, length, mass


# from this data frame, only choose proteins that have a HIGHER-THAN-AVERAGE LENGTH


# use the data frame above to make a scatter plot of length vs mass


# clean up/add labels


# change the look of your plot


```

# Exploring Esquisse

```{r}
# install the esquisse package from CRAN

# run esquisser() on the `proteins` data frame
# use this tool to make whatever plot you like
esquisse::esquisser(proteins)

# copy paste the esquisse code from above here
# go through the code line-by-line to make sure you understand what all of it does

```

