---
title: "Class Three (Answers): Computational Biology Reading Group"
author: "Your name"
date: "Nov 03, 2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("DESeq2"))
    BiocManager::install("DESeq2")
BiocManager::install("UCLouvain-CBIO/rWSBIM2122")

library("rWSBIM2122")
library(DESeq2)
library(tidyverse)
```

# Performing RNA-seq analysis using the DESeq2 package

For this walk-through we will be using the same example (and much of the same code!) as in this chapter - https://uclouvain-cbio.github.io/WSBIM2122/sec-rnaseq.html

# Load data 

```{r load}
# system.file() finds the location of the specified file in the specified package
# load() then opens the specified file (`.rda` objects in this case) so that they appear in our global environment 
load(system.file("extdata/deseq2/counts.rda",
                  package = "rWSBIM2122"))
load(system.file("extdata/deseq2/coldata.rda",
                  package = "rWSBIM2122"))
# you should now see two new objects in your Environment pane (top right-hand corner)

# `coldata` is a small dataframe, so just type its name in the console and press enter to see its contents
coldata

# `counts` is a much larger object. How many rows and columns does it have?
dim(counts)

# Take a look at the first few rows of `counts`
head(counts)

# Looking at these two dataframes, what kind of analysis/comparison would you perform?
```

# Construct a DESeqDataSet object
```{r dds}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design = ~ Condition) # what's this `design`? More about this later on

# How do we learn more about this new function, DESeqDataSetFromMatrix() 
?DESeqDataSetFromMatrix

# take a look at this new `dds` object
dds

# `dds` is a new type formally known as an "S4 object"
# S4 objects have slots that can be accessed by using the `@` symbol
# to see our two original objects `coldata` and `counts`, we could run the following
dds@colData
dds@assays@data[[1]]

# There are functions that allow us to access elements of `dds` in a more intuitive way
# use the `counts()` function to access the original `counts` dataframe
counts(dds)
# we can also use the `assay()` function for this

# use the `colData()` function to access the original (almost!) `coldata` dataframe
colData(dds)
# Is the output from above identical to the original `coldata` dataframe?

# subsetting `dds`
# only choose samples from the "mock" condition group
dds_mock <- dds[ , dds$Condition == "mock"]

# confirm that you only have "mock" samples
colData(dds_mock)

# take a look at the counts matrix for this new object. How many samples does this matrix have?
counts(dds_mock)

# using functions from the `tidyverse` package on `dds`
# plot the counts distribution of sample1

counts(dds) %>% 
  as_tibble() %>% 
  select(sample1) %>% 
  ggplot(aes(log2(sample1 + 1))) +
  geom_histogram(bins = 20)

# how would you plot the count distributions of ALL the samples?
# one quick approach is to use the base R `boxplot()` function
boxplot(log2(counts(dds) + 1))

# The tidyverse approach (boxplots or histograms) is a little more complicated...
counts(dds) %>% 
  as_tibble() %>% 
  pivot_longer(cols = everything(),
               names_to = "sample",
               values_to = "counts") %>% 
  ggplot(aes(log2(counts + 1), fill = sample)) +
  geom_histogram(bins = 20) +
  facet_wrap(~sample)
```

# Run DESeq2
```{r deseq2}
dds <- DESeq(dds)
# this function does a lot!
# let's ignore all of this for now and take a look at the results :)
results(dds)

# Notice that the log2FoldChange has the "KD" group as the base comparison
# why is this?
dds$Condition
class(dds$Condition)

# `Condition` is a factor, meaning that it is a categorical variable, which is a variable that contains a fixed number of categories or groups that a given observation can belong to
# The number of levels of a given factor refer to the number of categories this variable contains
# The order of these levels determines how comparisons are made in statistical modeling
# In the case of `Condition`, "KD" appears first and so is the base comparison group
# To switch this order, we can run this - 
res <- results(dds,
               contrast = c("Condition", "KD", "mock"))
res_df <- as_tibble(res, rownames = "ENSEMBL") # if we forget `rownames = `, we lose all the gene names!

# How would you remove all the rows that have `NAs` for the `padj` column
# We have a couple of options
# clue -
is.na(res_df$padj) # gives us a bunch of TRUEs and FALSEs. We can use this to filter out the unwanted rows
res_df_nona <- res_df %>% 
                filter(!is.na(padj))

# or we could use the `drop_na()` function
res_df_nona <- res_df %>% 
                  drop_na(padj)

# how to verify that there are no missing values for `padj`?
sum(is.na(res_df_nona$padj)) # this should be 0!
```

# Linear models, design formulae, and t-tests
```{r lm}
# make a dummy dataframe
lm_df <- tibble(
  condition = factor(c("control", "control", "control", "treated", "treated", "treated"),
                     levels = c("control", "treated")),
  reading = c(1.2, 1.6, 0.9, 3.4, 3.1, 2.4)
)

# confirm that the level order is correct for `condition`
levels(lm_df$condition)

# compare the two groups - "control" and "treated" using a t-test
t.test(reading ~ condition, data = lm_df, var.equal = TRUE)

# perform simple linear regression
lm_output <- lm(reading ~ condition, data = lm_df)
summary(lm_output)

# t.test and simple linear regression give us the same output!
```

For more on how common statistical tests are all just linear models in disguise, refer to this amazing article - 
https://lindeloev.github.io/tests-as-linear/#51_independent_t-test_and_mann-whitney_u

Next up - 
- Diving deeper into DESeq2
- PCA and clustering
- Pathway analysis
