---
title: "Class Four: Computational Biology Reading Group"
author: "Your name"
date: "Nov 10, 2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install CRAN packages
installifnot <- function(pckgName){
  if (!(require(pckgName, character.only = TRUE))) {
    install.packages(pckgName, dep = TRUE)
    require(pckgName, character.only = TRUE)
  }
}

cran_pkgs <- c("BiocManager", "remotes", "tidyverse", "pheatmap")

for (i in 1:length(cran_pkgs)){
  installifnot(cran_pkgs[i])
}

# Install Bioconductor packages
installBiocifnot <- function(pckgName){
  if (!(require(pckgName, character.only = TRUE))) {
    BiocManager::install(pckgName)
    require(pckgName, character.only = TRUE)
  }
}

bioc_pkgs <- c("DESeq2", "UCLouvain-CBIO/rWSBIM2122", "genefilter", "biomaRt")

for (i in 1:length(bioc_pkgs)){
  installBiocifnot(bioc_pkgs[i])
}

# Load packages
library(rWSBIM2122)
library(DESeq2)
library(tidyverse)
```

# Performing RNA-seq analysis using the DESeq2 package - continued

For this walk-through we will be using the same example (and much of the same code!) as in this chapter - https://uclouvain-cbio.github.io/WSBIM2122/sec-rnaseq.html

We will also be using the following vignettes as guides - 
For DESeq2 - http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#exploratory-analysis-and-visualization
For biomaRt - http://127.0.0.1:20249/library/biomaRt/doc/accessing_ensembl.html

# Load data and perform differential expression analysis

```{r load-and-run}
# system.file() finds the location of the specified file in the specified package
# load() then opens the specified file (`.rda` objects in this case) so that they appear in our global environment 
load(system.file("extdata/deseq2/counts.rda",
                  package = "rWSBIM2122"))
load(system.file("extdata/deseq2/coldata.rda",
                  package = "rWSBIM2122"))
# you should now see two new objects in your Environment pane (top right-hand corner)

# Construct a DESeqDataSet object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design = ~ Condition) 

# Run DESeq2
dds <- DESeq(dds)

res <- results(dds,
               contrast = c("Condition", "KD", "mock")) # we do this so that `mock` becomes the group we're comparing to
```

# Exploring the results table

```{r res}
# use `mcols` to learn more about the columns of `res`
mcols(res)

# to see the full descriptions, run this - 
mcols(res)$description

# PCA plot
# we need to first apply a regularized log transformation
# if we don't do this, the PCA will be dominated by genes with highest ABSOLUTE count
# this is because for RNA-seq count data, the variance increases with the mean 

rld <- rlogTransformation(dds)
plotPCA(rld)

# Calculate sample distances
sampleDists <- dist(t(assay(rld)))
sampleDists

# Plot heatmap of samples
# load the `pheatmap` package


# convert `sampleDists` to a matrix. Call this new matrix object `sampleDistMatrix`


# change the row names of `sampleDistMatrix` to be equal to the entries of `Condition`
 <- dds$Condition

# remove the column names of `sampleDistMatrix`
 <- NULL

# plot heatmap of samples
pheatmap(sampleDistMatrix)

# Plot heatmap of genes
library("genefilter")

# The `genefilter` package contains the helpful `rowVars()` function
# Use this function to pick the top 20 most variable genes

topVarGenes <- head(order(rowVars(assay(rld)), decreasing = TRUE), 20)
# what do the numbers in `topVarGenes` mean?
topVarGenes
# these numbers refer to the indices of the top 20 most variable genes in `rld`
# use these indices to create a matrix of the 20 most variable genes
mat <- 

# subtract the mean of each row from all the values in that row
# hint - use `rowMeans()`
mat <- 

# create a dataframe that contains the values of `Condition`
anno <- 
# clean up the column name of `anno`

# make sure the rownames of `anno` match the colnames of `mat`

anno

# plot the heatmap
pheatmap(mat, annotation_col = anno)
# compare with pheatmao(mat) to see what `annotation_col = anno` does

# Volcano plot
# Adding gene names
res_df <- as_tibble(res, rownames = "ENSEMBL") # if we forget `rownames = `, we lose all the gene names!
# Remove rows that have `NAs` in the `padj` column


# create a new column called neg_log_10_padj in `res_df`


# create a new column called `significant` that contain a TRUE if the following condition is satisfied - 
# padj < 0.05 & abs(log2FoldChange) > 1
# and contains a FALSE if the condition is NOT satisfied



# NOTE: all of the above can be achieved more efficiently by running the following
res_df %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = padj < 0.05 & abs(log2FoldChange) > 1)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(0.05)) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  theme(legend.position = "none")

# Get gene names
library(biomaRt)

# open vignettes
browseVignettes("biomaRt") # we will use the first vignette as a guide for the code below

# Step One: Connect to the selected BioMart database and dataset hosted by Ensembl
ensembl <- useEnsembl(biomart = "genes", 
                   dataset = "hsapiens_gene_ensembl")
# how do we know what databases and datasets are available?
# databases - 
listEnsembl()

# to see the datasets within the selected database - 
listDatasets(ensembl)

# Step Two: Retrieve gene names
# build a biomaRt query
# The getBM() function is the main query function in biomaRt
gene_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                 values = ,
                 mart = ensembl)

# how many missing gene names are there in `gene_names`


# add these gene names to res
res_df <- res_df %>% 
  left_join(gene_names, by = c("ensembl" = "ensembl_gene_id"))
```
To see how a `left_join()` works - 
https://github.com/gadenbuie/tidyexplain/blob/main/images/left-join.gif

Next up - 
- Enrichment Analyses (https://uclouvain-cbio.github.io/WSBIM2122/sec-gsea.html)

