---
title: "How to Perform RNA-seq analysis"
author: "Your name"
date: "Dec 01, 2021"

output:
  html_document: 
    
    toc: true 
    toc_float: 
      collapsed: false 
      smooth_scroll: false 
      
    number_sections: true 
    
    theme: paper 
    
    output: 
      fig_width: 10 
      fig_height: 8 
      fig_caption: true 
      dev: svg 
    
    #Dataframe printing
    df_print: paged 
    
    code_folding: show 
  
  pdf_document: default
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install CRAN packages
installifnot <- function(pckgName){
  if (!(require(pckgName, character.only = TRUE))) {
    install.packages(pckgName, dep = TRUE)
    require(pckgName, character.only = TRUE)
  }
}

cran_pkgs <- c("BiocManager", "remotes", "tidyverse", "plotly")

for (i in 1:length(cran_pkgs)){
  installifnot(cran_pkgs[i])
}

# Install Bioconductor packages
installBiocifnot <- function(pckgName){
  if (!(require(pckgName, character.only = TRUE))) {
    BiocManager::install(pckgName)
    require(pckgName, character.only = TRUE)
  }
}

bioc_pkgs <- c("DESeq2", "UCLouvain-CBIO/rWSBIM2122", "biomaRt", "org.Hs.eg.db", "clusterProfiler", "enrichplot")

for (i in 1:length(bioc_pkgs)){
  installBiocifnot(bioc_pkgs[i])
}

# Load packages
library(rWSBIM2122)
library(DESeq2)
library(tidyverse)
library(biomaRt)
library(org.Hs.eg.db)
library(clusterProfiler)
library(enrichplot)
library(plotly)
```

# Performing RNA-seq analysis using the DESeq2 package

For this walk-through we will be using the same example (and much of the same code!) as in this chapter - https://uclouvain-cbio.github.io/WSBIM2122/sec-rnaseq.html

For a basic intro to R and `dplyr`, please watch this series of videos made by Duke's Center for Computational Thinking - https://warpwire.duke.edu/w/f0YGAA/

To learn more about the DESeq2 package, please refer to this wonderful guide - http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#exploratory-analysis-and-visualization

# Load and Inspect Data 

```{r load}
# system.file() finds the location of the specified file in the specified package
# load() then opens the specified file (`.rda` objects in this case) so that they appear in our global environment 
load(system.file("extdata/deseq2/counts.rda",
                  package = "rWSBIM2122"))
load(system.file("extdata/deseq2/coldata.rda",
                  package = "rWSBIM2122"))
# you should now see two new objects in your Environment pane (top right-hand corner)

# `coldata` is a small dataframe, so just type its name in the console and press enter to see its contents
coldata

# `counts` is a much larger object. How many rows and columns does it have?
dim(counts)

# Take a look at the first few rows of `counts`
head(counts)

# Looking at these two dataframes, what kind of analysis/comparison would you perform?
```

# Construct a DESeqDataSet object

```{r dds}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design = ~ Condition)

# How do we learn more about this new function, DESeqDataSetFromMatrix() from the DESeq2 package
?DESeq2::DESeqDataSetFromMatrix

# take a look at this new `dds` object
dds

# `dds` is a new type formally known as an "S4 object"
# S4 objects have slots that can be accessed by using the `@` symbol
# to see our two original objects `coldata` and `counts`, we could run the following
dds@colData
head(dds@assays@data[[1]])

# There are functions that allow us to access elements of `dds` in a more intuitive way
# use the `counts()` function to access the original `counts` dataframe
head(counts(dds))
# we can also use the `assay()` function for this

# use the `colData()` function to access the original (almost!) `coldata` dataframe
colData(dds)
```

# Run DESeq2

## Run DESeq and inspect results
```{r deseq2}
dds <- DESeq(dds)
# this function does a lot!
# let's ignore all of this for now and take a look at the results :)
head(results(dds))
```

## Re-order levels and re-run DEseq
```{r re-order}
# Notice that the log2FoldChange has the "KD" group as the base comparison
# why is this?
dds$Condition
class(dds$Condition)

# `Condition` is a factor, meaning that it is a categorical variable, which is a variable that contains a fixed number of categories or groups that a given observation can belong to
# The number of levels of a given factor refer to the number of categories this variable contains
# The order of these levels determines how comparisons are made in statistical modeling
# In the case of `Condition`, "KD" appears first and so is the base comparison group
# To switch this order, we can run this - 
coldata$Condition <- factor(coldata$Condition, levels = c("mock", "KD"))

# check levels
coldata$Condition
# levels have been switched!

# create dds object again
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design = ~ Condition) 

# run DESeq again
dds <- DESeq(dds)
res <- results(dds)
head(res)
```

## Visualize with MA-plot
```{r ma-plot}
# visualize results using plotMA()
plotMA(res)
```

## Shrink log fold-changes for low-count genes
```{r lfcshrink}
res_shrunk <- lfcShrink(dds, 
                        coef = "Condition_KD_vs_mock",
                        type = "apeglm") 
# Here's a nice article on empirical Bayes estimation - http://varianceexplained.org/r/empirical_bayes_baseball/

plotMA(res_shrunk)
# compare this plot with the previous one. How are they different?
```

## Convert to dataframe and clean results table
```{r conv-and-clean}
# convert to dataframe
res_df <- as_tibble(res_shrunk, rownames = "ENSEMBL") # if we forget `rownames = `, we lose all the gene names!

# How would you remove all the rows that have `NAs` for the `padj` column
# We have a couple of options
# clue -
head(is.na(res_df$padj)) # gives us a bunch of TRUEs and FALSEs. We can use this to filter out the unwanted rows
res_df_nona <- res_df %>% 
                filter(!is.na(padj))

# or we could use the `drop_na()` function
res_df_nona <- res_df %>% 
                  drop_na(padj)

# how to verify that there are no missing values for `padj`?
sum(is.na(res_df_nona$padj)) # this should be 0!

res_df <- res_df_nona
```

## Make a Volcano Plot

To learn how to use ggplot2, refer to this video - https://www.youtube.com/watch?v=WUwSVKasU9g
```{r volcano}
res_df %>%
  ggplot(aes(x = log2FoldChange, y = -log10(padj),
             color = padj < 0.05 & abs(log2FoldChange) > 1)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(0.05)) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  theme(legend.position = "none")
```

# Over-representation Analysis (ORA)

For a more detailed explanation of enrichment analyses (including ORA and gene set enrichment analysis), please watch this video - https://youtu.be/ZgZKmAYm-LE

## Retrieve Entrez IDs and Gene Symbols
```{r entrez}
# Get ENTREZ IDs
# We'll need this for using the "org.Hs.eg.db" package
# Step One: Connect to the selected BioMart database and dataset hosted by Ensembl
ensembl <- useEnsembl(biomart = "genes", 
                   dataset = "hsapiens_gene_ensembl")

# Step Two: Retrieve gene names
# build a biomaRt query
# The getBM() function is the main query function in biomaRt
ensembl_to_entrez <- getBM(attributes = c("ensembl_gene_id", "external_gene_name",
                                            "entrezgene_id"),
                             values = res_df$ENSEMBL,
                             mart = ensembl)

# Plan B in case there's a connection problem -
# ensembl_to_entrez <- read_csv("https://raw.githubusercontent.com/dukecct/CBRG/main/inst/data/ensembl_to_entrez.csv")

# add this new info to res_df_nona
res_df <- res_df %>% 
            left_join(ensembl_to_entrez, by = c("ENSEMBL" = "ensembl_gene_id"))
```

## Clean `res_df`
```{r res-df-clean}
# remove rows with NAs in the columns in `entrezgene_id` and `padj`
res_df <- res_df %>% 
            drop_na(entrezgene_id, padj)

# are all ENTREZ IDs unique?
length(unique(res_df$entrezgene_id)) # 13711
nrow(res_df) # 13730

# drop duplicates
res_df <- res_df %>% 
            arrange(padj) %>% 
            distinct(entrezgene_id, .keep_all = TRUE)

# are all ENTREZ IDs unique?
length(unique(res_df$entrezgene_id)) == nrow(res_df) # TRUE
```

## Perform ORA on desired subset
```{r ora}
# perform ORA
# we need a vector of ENTREZ IDs for genes with padj values < 0.05
sig_genes <- res_df %>% 
              filter(padj < 0.05, log2FoldChange > 1) %>% 
              pull(entrezgene_id)
head(sig_genes)

go_ora <- enrichGO(gene = as.character(sig_genes),
                   OrgDb = org.Hs.eg.db,
                   universe = as.character(res_df$entrezgene_id),
                   ont = "MF",
                   readable = TRUE) # maps gene IDs to gene names
head(go_ora)

# Visualization
# dot plot
go_ora %>% 
  dotplot(showCategory = 30) + # showCategory	= number of enriched terms to display
  ggtitle("dotplot for ORA")

# or make your own plot!
ora_plot_interactive <- go_ora@result %>%
  dplyr::slice(1L:20L) %>% # show the first 20 enriched terms
  dplyr::mutate(GeneRatio = sapply(GeneRatio, function(x) eval(parse(text = x)))) %>% # compute decimal ratios) %>%
  # NOTE: Here we have switch from the pipe (%>%) to the "+" sign because ggplot2 only uses "+" (so far...)
  ggplot(aes(x = reorder(ID, GeneRatio), 
             y = GeneRatio, 
             fill = p.adjust,
             label = Description)) +
  geom_col() +
  coord_flip() +
  labs(x = "Enriched Term",
       y = "Gene Ratio") +
  theme_minimal()

plotly::ggplotly(ora_plot_interactive) # plotly CRAN package allows us to create interactive ggplot2 plots very easy!
```

