---
title: "Computational Biology Reading Group"
subtitle: "<br/>CBRG"
author: ""
institute: "DMPI"
date: "April 20, 2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
library(xaringanExtra)
library(janitor)
library(xaringanthemer)
library(pheatmap)
library(ggrepel)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
style_mono_accent(
  base_color = "#00539B",
  header_font_google = google_font("Adobe Caslon"),
  text_font_google   = google_font("Adobe Caslon", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r xaringanExtra-freezeframe, echo = FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
xaringanExtra::use_freezeframe()
```

```{r xaringanExtra-clipboard, echo=FALSE}
xaringanExtra::use_clipboard()
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: center, middle

# Hierarchical Clustering and Heatmaps 

---
# Example Dataset

```{r, echo = FALSE}
df <- tibble(
  gene = c("Gene A", "Gene B", "Gene C", "Gene D", "Gene E", "Gene F"),
  sample_one = c(1, 1.2, 0.9, 3.1, 2.9, 3.2),
  sample_two = c(0.9, 0.9, 0.8, 3, 3.1, 3.2)
)

df %>% 
  kbl() %>% 
  kable_classic()
```
---
.center[
```{r, echo = FALSE}
df %>% 
  ggplot(aes(sample_one, sample_two, label = gene)) +
  geom_point() +
  geom_text_repel() +
  theme_bw()
```
]
---
.center[
```{r, echo = FALSE}
df_dist <- dist(as.matrix(df[ , -1]))

plot(hclust(df_dist), labels = df$gene)
```
]
---
# Linkage Methods
.center[
```{r, echo = FALSE}
tibble(
  Linkage = c("Complete", "Single", "Average", "Centroid"),
  Description = c("Compute all pairwise dissimilairites between the observations in cluster A and the observations in cluster B, and record the largest of the similarities",
                  "...record the smallest of the similarities",
                  "...record the average of the similairites",
                  "Dissimilarity between the centroid for cluster A (a mean vector of length p) and the centroid for cluster B")
) %>% 
  kbl() %>% 
  kable_classic()
```
]
.footnote[Gareth James, Daniela Witten, Trevor Hastie, Robert Tibshirani. (2013). An introduction to statistical learning : with applications in R. New York :Springer]

---
# The Importance of Scaling
.center[
```{r, echo = FALSE}
df_2 <- tibble(
  gene = c("Gene A", "Gene B", "Gene C", "Gene D", "Gene E", "Gene F"),
  sample_one = c(1, 1.2, 0.9, 3.1, 2.9, 320),
  sample_two = c(2, 2.4, 1.8, 6.2, 5.8, 640)
)

df_2 %>% 
  kbl() %>% 
  kable_classic()
```
]
---
# Heatmap without scaling
.center[
```{r, echo = FALSE}
df_heat <- as.matrix(df_2[ , -1])
rownames(df_heat) <- df_2$gene

pheatmap(df_heat)
```
]
---
# Heatmap with scaling
.center[
```{r, echo = FALSE}
df_heat <- as.matrix(df_2[ , -1])
rownames(df_heat) <- df_2$gene

pheatmap(df_heat,
         scale = "row")
```
]
